Inference using context-dependent models of mutation

### 2021-10-13 writeups/context-dependence.tex
Finishing pass through main text

Fitting a model to nucleotide data would likely require iterative modification of the model.
The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data,
and $E_{x,y}$ for the number expected under the model.
(The expected number is the number of occurrences of $x$ in the initial sequence
multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

@@

For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
for various $t$.
This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
\begin{align}
    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
\end{align}
where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

Using the R package expm \citep{R_expm}, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
(and much longer for the whole matrix $e^{t G(8)}$).

@@

In phylogenetic applications, rather than ``before'' and ``after'' observations,
we get two (or more) observations evolved from a common root.
In the simplest case of two tips we have two processes $X$ and $Y$,
with identical starting states $X(0)=Y(0)$,
observed only at times $t_X$ and $t_Y$ respectively.

@@

The general case is a modification of Felsenstein's ``peeling'' algorithm.

@@

\end{enumerate}
i.e., only computing the likelihood for cases with $y=z=w$.
On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions
might be a significant loss of information,
but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)

@@

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time,
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the mcmc package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 60 independently simulated sequences,
and show the resulting credible intervals in Figure \ref{fig:ising_coverage}.
Poseterior medians are within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
As seen in Table \ref{tab:ising_coverage},
credible intervals were somewhat too narrow for shorter T-mers,
but $(6,2,2)$ T-mers were long enough with sufficient overhang to produce well-calibrated posteriors.

@@

\paragraph{Limitations}
The model presented assumes that the same process occurs at every location along the sequence,
and that the length of the sequnce does not change (no indels).
Neither of these assumptions are likely to be true for any data set of DNA sequence;
however, both assumptions are also made the state of the art in phylogenetics.
Some relaxation of the assumptions is possible -- for instnace,
we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
Short deletions might be accomodated by introducing a ``gap'' letter to the alphabet,
but insertions may be problematic.

\paragraph{Computation}
The main factor that determines computation time
is the lengths of the T-mers used for inference,
since all computations scale with the number of possible patterns at the long end of the T-mer.
Although the model is efficent enough to deal with fairly long T-mers,
this exponential growth puts hard constraints on the application of the method.
The length of T-mers required to accurately compute the likelihood in most models grows at most linearly
with the mean density of substitution between the initial and final sequence.
This does not in practice greatly increase the length of required T-mer
since substitution densities approaching 100\% are unlikely to be usable in practice
especially if how the sequences align with each other is unknown.
A more serious determining factor of T-mer length is the degree of dependency in the model itself.
For instance, inserting the 13-base pair motif selected against by meiotic drive in primates \citep{myers2010drive}
would require T-mers of width at least 25, and hence vectors of length $4^25 \approx 1.13 \times 10^15$.

\paragraph{Modeling}
Careful building of complex models in this framework may take care,
but no more than in any other class of models (e.g., multivariate linear regression).
It is easy to specify nonidentifiable models --
for instance, all 2mer mutation motifs include the 1mer motifs.
The large number of possible mutation motifs also makes multiple testing an issue;
an unguided approach might include all possible motifs but place a shrinkage prior on their values \citep{bhadra2015horseshoe}.
The converse approach of adding motifs that are overrepresented in the residuals
may be quicker, but entails some degree of arbitrariness in choosing the motifs to add (and in which order).
Finally, inference on trees may be difficult unless the root placement and distribution are strongly constrained:
otherwise, initial investigations suggest the likelihood surface is often strongly ridged --
for instance, increasing the abundance of \nG nucleotides at the root while also increasing the mutation rate away from \nG
can achieve roughly the same result.
However, in phylogenetic applications outgroups can be used to constrain these quantities.

>>>>

Fitting a model to nucleotide data would likely require iterative modification of the model.
The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data,
and $E_{x,y}$ for the number expected under the model.
(The expected number is the number of occurrences of $x$ in the initial sequence
multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
%EM I think n is used here in a different sense than just above.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

%EM This is beautifully simple but I'm not used to seeing p-values as a means of doing model selection (this is my lacking stats background showing up again).
%EM Can you point me to other examples where this is done?

@@

%EM This needs some notation updating. I think we want to commit to the n, \ell, r notation just above? I can do this.
For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
for various $t$.
This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
\begin{align}
    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
\end{align}
where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

%EM Is it worth noting that (IIUC) we could have additional sparsity based on the underlying (not T-mer) model? E.g. when you are are doing stepwise addition of parts to your model, you aren't using all possible 2m+1 - mers?
Using the R package expm \citep{R_expm}, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
(and much longer for the whole matrix $e^{t G(8)}$).

%EM What if we want to do derivative-based optimization of t? Then we'll have sparse times exp(sparse) times U. Still seems fast?

@@

%EM I skipped for now.
In phylogenetic applications, rather than ``before'' and ``after'' observations,
we get two (or more) observations evolved from a common root.
In the simplest case of two tips we have two processes $X$ and $Y$,
with identical starting states $X(0)=Y(0)$,
observed only at times $t_X$ and $t_Y$ respectively.

@@

%EM Peeling vs pruning?
The general case is a modification of Felsenstein's ``peeling'' algorithm.

@@

\end{enumerate}
i.e., only computing the likelihood for cases with $y=z=w$.
On a three-taxon tree it is not clear that this is desirable, since missing out on double transitions
might be a significant loss of information,
but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)

@@

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time,
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the mcmc package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 60 independently simulated sequences,
and show the resulting credible intervals in Figure~\ref{fig:ising_coverage}.

Posterior medians were within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
As seen in Table~\ref{tab:ising_coverage},
%EM I think this credible interval statement is interesting. So it's definitely this narrowness vs just having the estimate be off?
credible intervals were somewhat too narrow for shorter T-mers,
but $(6,2,2)$ T-mers were long enough with sufficient overhang to produce well-calibrated posteriors.

@@

\paragraph{Limitations}
%EM This is true for the implementation, but it's not clear that we couldn't do some of the same tricks for calculating likelihoods if we were handed a model.
The model presented assumes that the same process occurs at every location along the sequence,
and that the length of the sequnce does not change (no indels).
Neither of these assumptions are likely to be true for any data set of DNA sequence;
however, both assumptions are also made the state of the art in phylogenetics.
Some relaxation of the assumptions is possible -- for instnace,
we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
Short deletions might be accomodated by introducing a ``gap'' letter to the alphabet,
but insertions may be problematic.

\paragraph{Computation}
The main factor that determines computation time
is the lengths of the T-mers used for inference,
since all computations scale with the number of possible patterns at the long end of the T-mer.
Although the model is efficent enough to deal with fairly long T-mers,
this exponential growth puts hard constraints on the application of the method.
The length of T-mers required to accurately compute the likelihood in most models grows at most linearly
with the mean density of substitution between the initial and final sequence.
This does not in practice greatly increase the length of required T-mer
since substitution densities approaching 100\% are unlikely to be usable in practice
especially if how the sequences align with each other is unknown.
A more serious determining factor of T-mer length is the degree of dependency in the model itself.
For instance, inserting the 13-base pair motif selected against by meiotic drive in primates \citep{myers2010drive}
would require T-mers of width at least 25, and hence vectors of length $4^{25} \approx 1.13 \times 10^{15}$.

\paragraph{Modeling}
Careful building of complex models in this framework may take care,
but no more than in any other class of models (e.g., multivariate linear regression).
It is easy to specify nonidentifiable models --
for instance, all 2mer mutation motifs include the 1mer motifs.
The large number of possible mutation motifs also makes multiple testing an issue;
an unguided approach might include all possible motifs but place a shrinkage prior on their values \citep{bhadra2015horseshoe}.
The converse approach of adding motifs that are overrepresented in the residuals
may be quicker, but entails some degree of arbitrariness in choosing the motifs to add (and in which order).
Finally, inference on trees may be difficult unless the root placement and distribution are strongly constrained:
otherwise, initial investigations suggest the likelihood surface is often strongly ridged --
for instance, increasing the abundance of \nG\ nucleotides at the root while also increasing the mutation rate away from \nG\
can achieve roughly the same result.
However, in phylogenetic applications outgroups can be used to constrain these quantities.
